A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-15 latex/main.tex
Updates from Overleaf

We validated the recombination events in the sc2ts ARG by running 
the 3SEQ~\cite{Lam2018improved}, CovRecomb~\cite{Li2024CovRecomb} and rebar 
recombination detection methods on the putative recombinant sequences
(STAR Methods) summarised in Table~\ref{tab:recombinant_validation}.
Overall, validation among the QC-passing recombinants is highest
at with 3SEQ at 88\%, and is lowest for CovRecomb at xx\%. 
CovRecomb has been previously shown to be 
less sensitive than other methods~\cite{Li2024CovRecomb}, 
and is specifically intended 
to detect interlineage recombinants. The 12\% of recombinants 
not passing 3SEQ validation are also the most marginally supported
(see next section; Figure~\ref{fig:recombinant_evidence}).
All of the validation metrics also correlate strongly with the 
QC filter, with very few of the QC-failing recombinants
passing external validation (Table~\ref{tab:recombinant_validation}).

\subsection*{Assessing evidence for recombination events}
The LS HMM at the heart of sc2ts provides a simple and powerful means 
of quantifying the plausibility of a given recombination event.
To do this, we force the HMM to find a Viterbi solution for the 
putative recombinant sequence with no recombination by specifying
a large recombination penalty $k$ (Figure~\ref{fig:methodology}).
The difference between the numbers of mutations between
the two solutions---the number of mutations
``averted'' by recombination---is a measure of
the evidence in favor of recombination  (similar to 
idea of ``parsimony improvement'' in RIPPLES.)
Table~\ref{tab:pango_x_lineages} shows the number of averted mutations
for each of the 21 recombination events associated with Pango X lineages,
with support ranging from very strong in the case of XC
(the most parsimonious non-recombinant placement would
require 37 additional mutations) to marginal for XG and XW,
which have the minimal possible number of averted mutations.

@@

Recombination is rare in SARS-CoV-2, and ultimately judgements about
plausibility require expert analysis. 
To aid such interpretation and help identify quality control
issues we developed a visualisation 
based on SnpIt~\cite{OToole2024} and 
the recombinant view from RIVET~\cite{Smith2023tracking},
showing the informative genomic positions between 
a recombinant and its parent sequences, and highlighting 
potentially problematic loci.
Each copying pattern shows the positions
where the allelic state in a recombinant
(middle row) matches that of the left (upper row, coloured green) 
or right parent (lower row, coloured blue),
or where the recombination event requires a de-novo mutation
(gold, with mutational change below).
Figure~\ref{fig:recombinant_evidence}D shows some example copying patterns,
illustrating this spectrum of plausibility.
XA is an example of a highly plausible recombinant, shown by the 
large number of supporting loci on the left and right. As we have 
only one de-novo mutation, we can be sure that the parents are closely
observed [Some details about XA here, how well the parents were sampled etc].
XBB is also strongly supported by many loci on the left and right, 
but the presence of many de-novo mutations
observed in the copying patterns suggests the existence of
many genetically distinct, unsampled ancestors 
[FIXME: in this case, probably BJ.1, see issue 336],
and some caution should be exercised in interpreting them. 
See Figures~\ref{fig:recombinant_evidence}D and
\ref{fig:pango_x_copying_patterns} for copying patterns of the 
Pango X lineages, and 
Figure~\ref{fig:copying_pattern_examples} for copying patterns
highlighting specific types of QC problem (and more details on the 
visualisation).

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{inference-resources}
\caption{
Usage of computational resources
during the ARG inference process.
(A) The amount of time to process a daily batch of samples
peaks when the sample volume is the highest.
(B) Mean time to process a sample remained low.
(C) Max RAM usage also peaks around the time
that sampling effort is the highest.
}
\label{fig:inference_resources}
\end{figure}

>>>>

We validated the recombination events in the sc2ts ARG by running 
the 3SEQ~\cite{Lam2018improved}, CovRecomb~\cite{Li2024CovRecomb} and rebar 
recombination detection methods on the putative recombinant sequences
(STAR Methods) summarized in Table~\ref{tab:recombinant_validation}.
Overall, validation among the QC-passing recombinants is
the highest for 3SEQ at 88\% and
the lowest for CovRecomb at 25\%.
CovRecomb has been previously shown to be
less sensitive than other methods~\cite{Li2024CovRecomb}, 
and is specifically intended to detect interlineage recombinants.
As expected, CovRecomb missed most of the recombinants that
involve closely related parental Pango lineages
(Figure~\ref{fig:recombinant_evidence_cov_recomb}C),
in contrast to RIPPLES
which can detect intralineage recombinants
(see Figure~\ref{fig:recombinant_evidence_ripples_p4} and
Figure~\ref{fig:recombinant_evidence_ripples_p3}).
The 12\% of recombinants
not passing 3SEQ validation are also the most marginally supported
(see next section; Figure~\ref{fig:recombinant_evidence}).
All of the validation metrics also correlate strongly with the 
QC filter, with very few of the QC-failing recombinants
passing external validation (Table~\ref{tab:recombinant_validation}).

\subsection*{Assessing evidence for recombination events}

The LS HMM at the heart of sc2ts provides a simple and powerful means 
of quantifying the plausibility of a given recombination event.
To do this, we force the HMM to find a Viterbi solution for the 
putative recombinant sequence with no recombination by specifying
a large recombination penalty $k$ (Figure~\ref{fig:methodology}).
The difference between the numbers of mutations between
the two solutions---the number of mutations
``averted'' by recombination---is a measure of
the evidence in favor of recombination  (similar to 
idea of ``parsimony improvement'' in RIPPLES.)
Table~\ref{tab:pango_x_lineages} shows the number of averted mutations
for each of the 21 recombination events associated with Pango X lineages,
with support ranging from very strong in the case of XC
(the most parsimonious non-recombinant placement would
require 37 additional mutations) to marginal for XG and XW,
which have the minimal possible number of averted mutations.

@@

Recombination is rare in SARS-CoV-2, and
ultimately judgments about plausibility require expert analysis.
To aid such interpretation and help identify quality control issues
we developed a visualization tool
inspired by SnpIt~\cite{OToole2024} and 
the recombinant view of RIVET~\cite{Smith2023tracking},
showing the recombination-informative sites
between a recombinant and its parent sequences, and
highlighting potentially problematic loci.
Each copying pattern shows the positions
where the allelic state in a recombinant
(middle row) matches that of the left (upper row, colored green) 
or right parent (lower row, colored blue),
or where the recombination event requires a \emph{de novo} mutation
(gold, with mutational change below).
Figure~\ref{fig:recombinant_evidence}D shows some example copying patterns,
illustrating this spectrum of plausibility.
XA is an example of a highly plausible recombinant, shown by the 
large number of supporting loci on the left and right. As we have 
only one \emph{de novo} mutation, we can be sure that the parents are closely
observed [Some details about XA here, how well the parents were sampled etc].
XBB is also strongly supported by many loci on the left and right, 
but the presence of many \emph{de novo} mutations
observed in the copying patterns suggests the existence of
many genetically distinct, unsampled ancestors 
[FIXME: in this case, probably BJ.1, see issue 336],
and some caution should be exercised in interpreting them. 
See Figures~\ref{fig:recombinant_evidence}D and
\ref{fig:pango_x_copying_patterns}
for copying patterns of the Pango X lineages, and
Figure~\ref{fig:copying_pattern_examples} for copying patterns
highlighting specific types of QC problems
(and more details on the visualizations).

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{inference-resources}
\caption{
Usage of computational resources
during the ARG inference process.
(A) The amount of time to process a daily batch of samples
peaks when the sampling effort is the highest.
(B) Mean time to process a sample remained low throughout,
except towards the very end when there were few exact matches.
The HMM cost was moderate initially,
then noticeably decreased when sampling effort is the highest
before increasing again towards the end.
(C) Max RAM usage also peaks around the time
when sampling effort is the highest.
}
\label{fig:inference_resources}
\end{figure}
